Generalized Bradley-Terry models

### 2025-03-18 inst/jss_hyper3_real.Rnw
more suggestions from Gavin

where $w_{12}$ and $w_{34}$ are the number of wins by each team
respectively.  Then $\lambda$ would represent the effectiveness of
$p_1$'s special technique, and $\mu$ the enhancement afforded to $p_4$
by the drug; values $>1$ would correspond to a positive effect.  In
this paper I generalize the {\tt hyper2} R
package~\citep{hankin2017_rnw}, available on CRAN~\citep{rcore2025},
for Bradley-Terry likelihoods to accommodate this class of
multiplicative factors.  The most general likelihood function would be
proportional to

@@

The probability model of Figure~\ref{tikzabcde} is applied, in a
simple one-parameter form, to two datasets: one a set of preferences
of different types of sushi, and one in the field of esports.

\section{Red bus, blue bus}

@@

Here I suggest an alternative in the form of a generalization of
Plackett-Luce likelihood functions for ranking options, using the
multiplicative generalization to Bradley-Terry strengths presented
above.  If we ask our respondent to rank his options, it is highly
probable that he will put RB and BB consecutively (because they are
essentially indistinguishable).  Can we quantify the strength of this
effect?  To do this, we define a bespoke function \code{RB\_BB\_LF()}
which returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options:

<<defineredbuslf>>=
`RB_BB_LF` <- function(lambda){
    ec <- c(C=1, T=2, RB=3, BB=3, W=4) 
    h <- c(1, 1, lambda, 1)           
    (
        cheering3(v=c("RB", "BB", "C" , "T", "W"), e=ec, h=h)*3 + 
        cheering3(v=c("BB", "RB", "T" , "C", "W"), e=ec, h=h)*2 + 
        cheering3(v=c("T" , "BB", "RB", "C", "W"), e=ec, h=h)*5 + 
        cheering3(v=c("W" , "BB", "RB", "T", "C"), e=ec, h=h)*4 + 
        cheering3(v=c("C" , "RB", "BB", "W", "T"), e=ec, h=h)*3 + 
        cheering3(v=c("BB", "C" , "RB", "T", "W"), e=ec, h=h)*3
    )
}
@

@@

We now apply the ideas above to a dataset obtained from the {\tt
  preflib} library~\citep{MaWa13a} in which ten different types of
sushi were placed in order of preference by 5000 judges.  Three of the
sushis (maguro, toro, tekka-maki) were made with tuna, and it is
reasonable to view these three types as being exchangeable in much the
same way as the red and blue buses.  To proceed, we need to define
equivalence classes of sushi---one for tuna-based sushis and one for
everything else.  Of course, one can also imagine the opposite
phenomenon whereby the judge decides he has had enough tuna for the
moment and this disincentivises further tuna choices.  This mechanism
would be modelled by having $\lambda<1$.

Implementation in \code{hyper3} formalism is straightforward.  The
dataset used is \code{sushi\_table}, part of the \code{hyper2} package:

@@

Bradley-Terry strengths $p_1,\ldots,p_n$, $p_i\geqslant 0$, $\sum
p_i=1$ are generalized with non-negative multiplicative $\lambda$-
factors that represent temporary or ephemeral changes in competitive
strength.  The {\tt hyper2} R package implements the resulting
likelihood functions, and furnishes a wide range of functionality
geared towards assessing meaningful nulls using likelihood methods.

The standard Plackett-Luce likelihood function
$\prod_{i=1}^{n}p_{(i)}\left[\sum_{j=i}^n p_{(j)}\right]^{-1}$ over
Bradley-Terry strengths assumes conditional independence of successive
rank observations $p_{(1)},\ldots p_{(n)}$, an assumption which is
broken in the current work.  As indicated above, multiplicative
$\lambda$- factors may be used to implement many different
non-independence schemes and comparing these might be interesting.

>>>>

where $w_{12}$ and $w_{34}$ are the number of wins by each team
respectively.  Then $\lambda$ would represent the effectiveness of
$p_1$'s special technique, and $\mu$ the enhancement afforded to $p_4$
by the drug; values $>1$ would correspond to a positive effect.  In
this paper I generalize the {\tt hyper2} R
package~\citep{hankin2017_rnw}, available on CRAN~\citep{rcore2025},
for Bradley-Terry likelihoods to accommodate this class of
multiplicative factors with new \proglang{S3} class {\tt hyper3} and
associated methods.

The most general likelihood function would be
proportional to

@@

The probability model of Figure~\ref{tikzabcde} is applied, in a
simple one-parameter form, to two datasets: one a set of preferences
of different types of sushi, and one in the field of esports.  First,
however, we discuss a synthetic example showing how generalizing
Placket-Luce to admit non-independence of successive arrivals can be
applied to an apparently different situation.

\subsection{Red bus, blue bus}

@@

Here I suggest an alternative in the form of a one-parameter
generalization of Plackett-Luce likelihood functions for ranking
options, using the multiplicative generalization to Bradley-Terry
strengths presented above.  As a synthetic example, we specify options
RB and BB for red bus and blue bus together options C, T and W for
car, train, and walking.

Specifically, it seems reasonable to assume that if we ask our
respondent to rank his options, it is highly probable that he will put
RB and BB consecutively (because they are essentially
indistinguishable).  Can we quantify the strength of this effect?  To
do this, we define a bespoke function \code{RB\_BB\_LF()} which
returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options.

<<defineredbuslf>>=
`RB_BB_LF` <- function(lambda){
    ec <- c(C=1, T=2, RB=3, BB=3, W=4) 
    h <- c(1, 1, lambda, 1)           
    (
        cheering3(v=c("RB", "BB", "C" , "T", "W"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "RB", "T" , "C", "W"), e = ec, h = h) * 2 + 
        cheering3(v=c("T" , "BB", "RB", "C", "W"), e = ec, h = h) * 5 + 
        cheering3(v=c("W" , "BB", "RB", "T", "C"), e = ec, h = h) * 4 + 
        cheering3(v=c("C" , "RB", "BB", "W", "T"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "C" , "RB", "T", "W"), e = ec, h = h) * 3
    )
}
@

@@

We now apply the ideas above to a dataset obtained from the {\tt
  preflib} library~\citep{MaWa13a} in which ten different types of
sushi were placed in order of preference by 5000 judges.  Three of the
sushis (maguro, toro, tekka-maki) were made with tuna, and it is
reasonable to view these three types as being exchangeable in much the
same way as the red and blue buses.  To proceed, we need to define
equivalence classes of sushi---one for tuna-based sushis and one for
everything else.  Of course, one can also imagine the opposite
phenomenon whereby the judge decides he has had enough tuna for the
moment and this disincentivises further tuna choices.  This mechanism
would be modelled by having $\lambda<1$.  Implementation in
\code{hyper3} formalism is straightforward.  The dataset used is
\code{sushi\_table}, part of the \code{hyper2} package:

@@

Bradley-Terry strengths $p_1,\ldots,p_n$, $p_i\geqslant 0$, $\sum
p_i=1$ are generalized with non-negative multiplicative $\lambda$-
factors that represent temporary or ephemeral changes in competitive
strength.  The {\tt hyper2} R package implements the resulting
likelihood functions using \proglang{S3} class {\tt hyper3}, and
furnishes a wide range of functionality geared towards assessing
meaningful nulls using likelihood methods.

The standard Plackett-Luce likelihood function
$\prod_{i=1}^{n}p_{(i)}\left[\sum_{j=i}^n p_{(j)}\right]^{-1}$ over
Bradley-Terry strengths assumes conditional independence of successive
rank observations $p_{(1)},\ldots p_{(n)}$, an assumption which is
broken in the current work.  As indicated above, multiplicative
$\lambda$- factors may be used to implement many different
non-independence schemes.
